---
output: github_document
always_allow_html: true
---

# distantia: Time Series Dissimilarity Analysis with Dynamic Time Warping

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = TRUE
)
```

<!-- badges: start -->

[![DOI](https://zenodo.org/badge/187805264.svg)](https://zenodo.org/badge/latestdoi/187805264)
[![CRAN\_Release\_Badge](http://www.r-pkg.org/badges/version/distantia)](https://CRAN.R-project.org/package=distantia)
[![CRAN\_Download\_Badge](http://cranlogs.r-pkg.org/badges/grand-total/distantia)](https://CRAN.R-project.org/package=distantia)
[![R-CMD-check](https://github.com/BlasBenito/distantia/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/distantia/actions/workflows/R-CMD-check.yaml)


<!-- badges: end -->

## Warning

Version 2.0.0 of `distantia` is a full re-write from scratch and **will break existing workflows**. Please refer to the [Changelog](https://blasbenito.github.io/distantia/news/index.html) for details before updating.

## Summary

The `distantia` package has undergone a major overhaul to enhance performance and usability. Core functions are now implemented in C++ for faster execution and improved memory efficiency, with streamlined R wrappers to simplify user interaction. 

Function and argument names have been updated to offer a more intuitive experience. Additionally, most time series operations are now powered by the [zoo](https://cran.r-project.org/web/packages/zoo/index.html) library to provide a consistent and efficient data handling.

Version 2.0 introduces "time series lists" (`tsl`), which are lists of zoo objects. A comprehensive suite of functions named `tsl_...()` have been added to the pacakge to generate, resample, transform, analyze, and visualize univariate, multivariate, regular, or irregular time series. Many of these functions support a parallelized execution via the [future](https://future.futureverse.org/) package, and progress bars as provided by the package [progressr](https://progressr.futureverse.org/).

Finally, to further aid learning and experimentation, new example datasets from different disicplines and tools to simulate time series are also included.

## Citation

If you find this package useful, please cite it as:

*Blas M. Benito, H. John B. Birks (2020). distantia: an open-source toolset to quantify dissimilarity between multivariate ecological time-series. Ecography, 43(5), 660-667. doi: [10.1111/ecog.04895](https://nsojournals.onlinelibrary.wiley.com/doi/10.1111/ecog.04895).*

*Blas M. Benito (2024). distantia: A Toolset for Time Series Dissimilarity Analysis. R package version 2.0.0. url:  [https://blasbenito.github.io/distantia/](https://blasbenito.github.io/distantia/).*

## Main Improvements in Version 2.0.0

1. **High computational speed**: The logics for the time series dissimilarity analysis provided by `distantia()` and `distantia_importance()` are now implemented in a C++ backend designed for speed. This new feature, combined with a parallelization setup provided by the [future](https://future.futureverse.org/) makes dissimilarity analysis for large datasets blazing fast.
2. **Usage of zoo objects**: Individual time series in distantia are now [zoo](https://cran.r-project.org/package=zoo) objects, which support regular and irregular time series with various index types (e.g., Date, POSIXct), enable efficient handling and manipulation, and offer robust tools for alignment, merging, and filling missing values.
3. **Time Series Lists and associated toolset**: Groups of time series to compare are stored in Time Series Lists (TSL), which are lists of zoo objects. The package also comes with a complete toolset to create, manage, transform, aggregate, and visualize TSLs, with the objective of providing a smooth user experience.

## Install

The package `distantia` can be installed from CRAN.

```{r, eval = FALSE}
install.packages("distantia")
```


The development version can be installed from GitHub.

```{r, eval = FALSE}
remotes::install_github(
  repo = "blasbenito/distantia", 
  ref = "main"
  )
```


Previous versions are in the “archive_xxx” branches of the GitHub repository.

```{r, eval = FALSE}
remotes::install_github(
  repo = "blasbenito/distantia", 
  ref = "archive_v1.0.2"
  )
```

## Getting Started

This section provides a minimal example on how to use `distantia`. Please, check the **Articles** section for further details.

### Setup

The code below loads the required packages, and defines a parallelization backend.

```{r packages, message = FALSE, warning = FALSE}
#required packages
library(distantia)
library(future)
library(parallelly)
library(tmap)

#parallelization setup
future::plan(
  future::multisession,
  workers = parallelly::availableCores() - 1
  )

#progress bar (does not work in Rmarkdown)
#progressr::handlers(global = TRUE)
```

### Example data

The data frame `cities_temperature` contains long-term time series of monthly temperatures for 20 large cities.

```{r}
head(cities_temperature)
```

The sf data frame `cities_coordinates` contains city coordinates.

```{r}
cities_coordinates[, "name"]
```
### Converting to TSL

The function `tsl_initialize()` transforms the target data frame into a time series list.

```{r}
tsl <- tsl_initialize(
  x = cities_temperature,
  name_column = "name",
  time_column = "time"
)

head(lapply(tsl, class))
```
The functions `tsl_subset()` and `tsl_transform()` are used below to focus on a smaller subset of the the data to facilitate the explanations below, and to center and scale the data to facilitate the comparison between methods.

```{r}
tsl_three_cities <- tsl |> 
  tsl_subset(
    names = c(
      "London", 
      "Paris", 
      "Kinshasa"
    ),
    time = c("2000-01-01", "2005-01-01")
  ) |> 
  tsl_transform(
    f = f_scale #centering and scaling
  )
```

The function `tsl_time()` offers a quick look at the time features of the time series list.

```{r}
tsl_time(tsl = tsl_three_cities)[, c(
  "name", 
  "class", 
  "resolution", 
  "units", 
  "begin", 
  "end"
  )]
```
The function `tsl_plot()` facilitates a visual exploration of the data.

```{r}
tsl_plot(
  tsl = tsl_three_cities,
  ylim = "relative"
)
```


### Dissimilarity Analysis

The function `distantia()` offers two alternative methods to computes the dissimilarity between pairs of time series: 

 - **Dynamic Time Warping (DTW)** (default): Warps time axes to align sections with similar shapes. This method is ideal for comparing time series with temporal shifts and is applicable to any type of time series. However, DTW is sensitive to differences in magnitudes, so it performs best when the data is scaled, and optionally centered if absolute values are not critical for the analysis.       
 
 - **Lock-step**: Compares values at corresponding time points directly, without any temporal adjustments. This method is suitable for cases where preserving temporal integrity is essential, but it requires time series sampled at the same time points. This method does not require scaling unless the results are to be compared with DTW, in which case scaling ensures comparable results across both methods.


#### Dynamic Time Warping

When `lock_step = FALSE`, dynamic time warping is used to compare the time series.

```{r}
df_dtw <- distantia(
  tsl = tsl_three_cities,
  distance = "euclidean",
  lock_step = FALSE #default, activates dynamic time warping
)

df_dtw[, c("x", "y", "psi")]
```

The column "psi" represents the normalized dissimilarity between pairs of time series. According to these values, London and Kinshasa show the most dissimilar temperatures, while London and Paris show the most similar ones.

The time warping analysis can be plotted with `distantia_plot()`. The temperature patterns of "Paris" and "London" are quite similar, and require no warping.

```{r}
distantia_plot(
  tsl = tsl_three_cities[c("Paris", "London")]
)
```

The warping alignment between "London" and "Kinshasa" shows the adjustments made by DTW to match the peaks and valleys of both time series, and compensate for their shifted patterns.

```{r}
distantia_plot(
  tsl = tsl_three_cities[c("London", "Kinshasa")]
)
```

When the data is not scaled and/or centered, the sensitivity to data magnitude of DTW may lead to local optima that distort the warping. The example below comparing the raw temperatures of London and Kinshasa shows two of these local optima, one in the *y* axis, at the minimum temperature in Kinshasa (winter of 2001), and another in the *x* axis, at the maximum temperature in London (summer of 2003).

```{r}
tsl_two_cities <- tsl |> 
  tsl_subset(
    names = c(
      "London", 
      "Kinshasa"
    ),
    time = c("2000-01-01", "2005-01-01")
  )

distantia_plot(
  tsl = tsl_two_cities
)
```

This is not a particular issue of the algorithm implemented in `distantia`, but a general behavior of time warping methods. The code below shows a similar analysis performed with [`dtw`](https://CRAN.R-project.org/package=dtw).

```{r, message = FALSE}
library(dtw)

xy_dtw <- dtw::dtw(
  x = tsl_two_cities$London$temperature,
  y = tsl_two_cities$Kinshasa$temperature,
  keep = TRUE
  )

plot(xy_dtw, type = "threeway")
```

#### Lock Step

When `lock_step = TRUE`, the lock-step method is activated, and the time series are compared on an observation-to-observation basis. This method yields increased dissimilarity scores when the involved time series show shifted patterns.

```{r}
df_ls <- distantia(
  tsl = tsl_three_cities,
  distance = "euclidean",
  lock_step = TRUE #activates lock-step
)

#adding the results of dtw to facilitate comparison
df_ls$psi_dtw <- df_dtw$psi
df_ls$psi_lock_step <- df_ls$psi

df_ls[, c("x", "y", "psi_lock_step",  "psi_dtw")]
```
It is important to take in mind that here we are performing the lock-step analysis with the scaled data to compare its output with the DTW results. The results on the raw data are shown below.

```{r}
tsl_three_cities <- tsl |> 
  tsl_subset(
    names = c(
      "London", 
      "Paris", 
      "Kinshasa"
    ),
    time = c("2000-01-01", "2005-01-01")
  )

df_ls <- distantia(
  tsl = tsl_three_cities,
  distance = "euclidean",
  lock_step = TRUE #activates lock-step
)

df_ls[, c("x", "y", "psi")]
```


### Manipulating TSLs

This section only covers a few of the data-manipulation capabilities of `distantia`.

#### Aggregation

The function `tsl_aggregate()` applies an aggregation function to specific time frames. For example, the code below computes the maximum temperature per year

```{r}
tsl_temp <- tsl_aggregate(
  tsl = tsl,
  new_time = "year",
  method = max
)

tsl_plot(
  tsl = tsl_temp[1:5],
  ylim = "relative"
)
```
Same as above, but by decade.

```{r}
tsl_temp <- tsl_aggregate(
  tsl = tsl,
  new_time = "decades",
  method = max
)

tsl_plot(
  tsl = tsl_temp[1:5],
  ylim = "relative"
)
```

## Getting help

If you encounter bugs or issues with the documentation, please [file a issue on GitHub](https://github.com/BlasBenito/distantia/issues).
