---
output: github_document
always_allow_html: true
---

# distantia: Time Series Dissimilarity Analysis with Dynamic Time Warping

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = TRUE
)
```

<!-- badges: start -->

[![DOI](https://zenodo.org/badge/187805264.svg)](https://zenodo.org/badge/latestdoi/187805264)
[![CRAN\_Release\_Badge](http://www.r-pkg.org/badges/version/distantia)](https://CRAN.R-project.org/package=distantia)
[![CRAN\_Download\_Badge](http://cranlogs.r-pkg.org/badges/grand-total/distantia)](https://CRAN.R-project.org/package=distantia)
[![R-CMD-check](https://github.com/BlasBenito/distantia/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BlasBenito/distantia/actions/workflows/R-CMD-check.yaml)


<!-- badges: end -->

## Warning

Version 2.0.0 of `distantia` is a full re-write of the original package and **will break existing workflows**. Please refer to the [Changelog](https://blasbenito.github.io/distantia/news/index.html) for details before updating.

## Summary

The R package **`distantia`** offers an efficient, feature-rich toolkit for managing, comparing, and analyzing time series data. It is designed to handle a wide range of scenarios, including:

- Multivariate or univariate time series.
- Regular or irregular sampling. 
- Time series of different lengths.

### Key Features

- **Comprehensive Analytical Tools**:
  - 10 distance metrics: see `distantia::distances`.
  - The normalized dissimilarity metric `psi`.
  - Three Dynamic Time Warping (DTW) methods for shape-based comparisons.
  - A Lock-Step method for sample-to-sample alignment.
  - Restricted permutation tests for robust inferential support.
  - Variable Importance Analysis: assessment of contribution to dissimilarity of individual variables in multivariate time series.
  - Hierarchical and K-means clustering of time series based on dissimilarity matrices.
  
- **Computational Efficiency**:
  - A **C++ back-end** powered by [Rcpp](https://www.rcpp.org/).
  - **Parallel processing** managed through the [future](https://future.futureverse.org/) package.
  - **Efficient data handling** via [zoo](https://cran.r-project.org/web/packages/zoo/index.html).

- **Time Series Management Tools**:
  - Introduces **time series lists (`tsl`)**, a versatile format for handling collections of time series stored as lists of `zoo` objects.
  - Includes a suite of `tsl_...()` functions for generating, resampling, transforming, analyzing, and visualizing both univariate and multivariate time series.

## Citation

If you find this package useful, please cite it as:

*Blas M. Benito, H. John B. Birks (2020). distantia: an open-source toolset to quantify dissimilarity between multivariate ecological time-series. Ecography, 43(5), 660-667. doi: [10.1111/ecog.04895](https://nsojournals.onlinelibrary.wiley.com/doi/10.1111/ecog.04895).*

*Blas M. Benito (2024). distantia: A Toolset for Time Series Dissimilarity Analysis. R package version 2.0.0. url:  [https://blasbenito.github.io/distantia/](https://blasbenito.github.io/distantia/).*

## Install

The package `distantia` can be installed from CRAN.

```{r, eval = FALSE}
install.packages("distantia")
```


The development version can be installed from GitHub.

```{r, eval = FALSE}
remotes::install_github(
  repo = "blasbenito/distantia", 
  ref = "development"
  )
```

## Getting Started

This section provides a minimal example on how to use `distantia`. Please, check the **Articles** section for further details.

### Setup

Most functions in `distantia` now support a parallelization backend and progress bars.

```{r packages, message = FALSE, warning = FALSE}
#required packages
library(distantia)
library(future)
library(parallelly)

#parallelization setup
future::plan(
  future::multisession,
  workers = parallelly::availableCores() - 1
  )

#progress bar (does not work in Rmarkdown)
#progressr::handlers(global = TRUE)
```

### Example Data

The `albatross` data frame contains daily GPS data of 4 individuals of Waved Albatross in the Pacific captured during the summer of 2008. The first 10 rows of this data frame are shown below.

```{r, echo = FALSE}
albatross |> 
  distantia::utils_drop_geometry() |> 
  head()
```

The `albatross` data is converted to *Time Series List* with `tsl_initialize()`, and scaled and centered with `tsl_transform()` and `f_scale`, to facilitate shape-based comparisons and variable importance analyses.

```{r, fig.width=5, fig.height=4.5}
tsl <- tsl_initialize(
  x = albatross,
  name_column = "name",
  time_column = "time",
  lock_step = TRUE
) |> 
  tsl_transform(
    f = f_scale
  )

tsl_plot(
  tsl = tsl,
  ylim = "relative"
)
```

### Dissimilarity Analysis

#### Lock-Step Analysis

The lock-step analysis performs a direct sample-to-sample comparison without any time distortion. It requires time series of the same length observed at the same times.

```{r}
df_ls <- distantia(
  tsl = tsl,
  lock_step = TRUE
)

df_ls[, c("x", "y", "psi")]
```
The column "psi" contains the value of normalized dissimilarity, and is used to order the data frame from lower to higher dissimilarity (first row shows the most similar time series).

The function `distantia_boxplot()` helps to quickly identify these time series that are more different (top) or similar (bottom) to all others.

```{r, fig.width=5, fig.height=4.5}
distantia_boxplot(df = df_ls)
```

#### Dynamic Time Warping

The function `distantia()` computes the dissimilarity between pairs of time series using **dynamic time warping** (DTW) with weighted diagonals by default. This analysis is useful when the time series might be shifted for reasons not relevant to the analysis at hand, such as varying phenology, asynchronic behavior, or non-overlappig sampling periods

```{r}
df_dtw <- distantia(
  tsl = tsl
)

df_dtw[, c("x", "y", "psi")]
```

The function `distantia_plot()` provides a detailed insight on the alignment between a pair of time series resulting from dynamic time warping.

```{r, fig.width=6, fig.height=5}
distantia_plot(
  tsl = tsl[c("X132", "X153")]
)
```

Regions of the least cost path deviating from a perfect diagonal show adjustments made by the dynamic time warping algorithm to match the time series by shape rather than by time. 

Vertical or horizontal segments of the least cost path indicate that one sample of one time series is being paired with several samples of the other time. 

### P-values

The package `distantia` implements restricted permutation tests to help assess the significance of dissimilarity scores. It accepts different setups to support different data assumptions. 

For example, the configuration below re-arranges complete rows within 7 days blocks. This setup assumes a strong dependency between variables in the same row, and a strong dependency between observations that are close in time (same week). 

```{r}
df_dtw <- distantia(
  tsl = tsl,
  repetitions = 1000,
  permutation = "restricted_by_row",
  block_size = 7
)

df_dtw[, c("x", "y", "psi", "p_value")]
```
The p-values shown in the results represent the fraction of permutations that yield a psi score lower than the observed, and can be interpreted as the *strength of the similarity* between two time series. 

Given a significance level of interest (i.e. 0.05, but depends on the number of iterations), it can help separate pairs of time series that are similar from pairs of time series that are dissimilar.

## Getting help

If you encounter bugs or issues with the documentation, please [file a issue on GitHub](https://github.com/BlasBenito/distantia/issues).
