---
title: "The Psi Dissimilarity Metric"
output: 
  rmarkdown::html_document:
    toc: true
    toc_title: "Content"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.show = 'hold' #side by side images
)
```

```{r setup}
library(distantia)
```

# Summary

This article, based on [Benito and Birks 2020](https://doi.org/10.1111/ecog.04895), describes in detail the step-by-step computation of the **Psi dissimilarity metric** used in `distantia` to compare pairs of time series. 

# The Psi Dissimilarity Metric

The Psi Dissimilarity Metric ($\psi$ hereafter) measures distance between two time series in the range $[0, \infty)$, where 0 represents identical time series. It was first described in [Birks and Gordon 1985](https://onlinelibrary.wiley.com/doi/abs/10.1002/gea.3340010406) and later extended and implemented in the R package `distantia` in [Benito and Birks 2020](https://doi.org/10.1111/ecog.04895). 

In essence, $\psi$ is the sum of distances between all *relevant* pairs of samples of two time series, normalized by the sum of distances between consecutive samples within each time series, as expressed below:

$$\psi = \frac{AB_{between} - AB_{within}}{AB_{within}}$$


where $A$ and $B$ are two time series represented as matrices with the same number of columns, $AB_{between}$ is the distance between time series, and $AB_{within}$ is the normalization factor. 

Going deeper into the math notation, $AB_{between}$ is expressed as:

$$AB_{\text{between}} = 2 \times \sum_{i=1}^{m=n} D(A_i, B_i)$$
where $m$ and $n$ are the number of samples in $A$ and $B$, $i$ is the relevant sample index, $n$ is the total number of sample pairs, and $D$ is a distance function (e.g., Euclidean, Manhattan, or any other relevant distance metric for the values in $A$ and $B$).

On the other hand, $AB_{within}$ is represented as:

$$AB_{\text{within}} = \sum_{i=2}^{m} D(A_{i-1}, A_i) + \sum_{i=2}^{n} D(B_{i-1}, B_i) $$
where $D(A_{i-1}, A_i)$ and $D(B_{i-1}, B_i)$ represent the sum of distances between consecutive samples within $A$ and $B$, respectively.
