---
title: "distantia mini-demo"
author: "Blas M. Benito"
format: html
editor: visual
---

## Summary

This is a demo of the R package `distantia` to announce the release of the version 2.0 in Bluesky and Fosstodon.

### 1

Well, that took a while (one year of intermittent development), but the version 2.0 of the #Rstats package {distantia} (URL: <https://blasbenito.github.io/distantia/>) is finally on CRAN.

```{r}
install.packages(
  pkgs = "distantia",
  dependencies = TRUE,
  quiet = TRUE
  )

utils::packageVersion(pkg = "distantia")
```

## 2

{distantia} provides tools to load, transform, and compare time series of any kind (uni or multivariate, regular or irregular, sync or async), with a focus on

### 3

{distantia 2.0} provides:

-   C++ implementations of the time series comparison methods dynamic time warping and lock-step.

-   Restricted permutation tests and variable importance assessment.

3

```{r, fig.width=5, fig.height=4.1}
library(distantia, quietly = TRUE)
tsl <- tsl_initialize(
  x = distantia::albatross,
  name_column = "name",
  time_column = "time"
) |> 
  tsl_transform(
    f = f_scale_local
  ) |> 
  tsl_colnames_set(
    names = c("longitude", "latitude", "speed", "temperature", "heading")
  ) |> 
  tsl_names_set(
    names = c("Bird_1", "Bird_2", "Bird_3", "Bird_4")
  )

distantia_dtw_plot(
  tsl = tsl[1:2]
)
```
