---
title: "Comparison old and new versions: old version"
author: "Blas M. Benito"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Installing old package version

```{r, echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE}
library(here)
Rcpp::compileAttributes(here::here())
devtools::document()
devtools::install()
library(distantia)
```

#Preparing comparison data

```{r}
data(sequenceA)
data(sequenceB)

sequences <- prepareSequences(
  sequence.A = sequenceA,
  sequence.B = sequenceB,
  merge.mode = "overlap",
  if.empty.cases = "omit",
  transformation = "hellinger"
)
```


# Comparison

## Distance matrix

```{r}
old_dist_matrix <- distanceMatrix(
  sequences = sequences,
  method = "euclidean"
)
old_dist_matrix[[1]][1:5, 1:5]
old_dist_matrix[[1]][45:49, 28:32]
```

## Cost matrix


```{r}
old_cost_matrix <- leastCostMatrix(
  distance.matrix = old_dist_matrix,
  diagonal = FALSE
)
old_cost_matrix[[1]][1:5, 1:5]
old_cost_matrix[[1]][45:49, 28:32]
```

## Cost path

### Orthogonal

```{r}
#extracting least-cost path
old_path <- leastCostPath(
  distance.matrix = old_dist_matrix,
  least.cost.matrix = old_cost_matrix
  )

nrow(old_path[[1]])
old_path
sum(old_path$`A|B`$distance)
sum(old_path$`A|B`$cumulative.distance)
```

### Diagonal

```{r}
#extracting least-cost path
old_path_diagonal <- leastCostPath(
  distance.matrix = old_dist_matrix,
  least.cost.matrix = old_cost_matrix,
  diagonal = TRUE
  )

nrow(old_path_diagonal[[1]])
old_path
sum(old_path_diagonal$`A|B`$distance)
sum(old_path_diagonal$`A|B`$cumulative.distance)
```

## Remove blocks

The old version removes valid data from the path file.

```{r}
old_path_no_blocks <- leastCostPathNoBlocks(
    least.cost.path = old_path
    )
nrow(old_path_no_blocks[[1]])
sum(old_path_no_blocks[[1]]$distance)
sum(old_path_no_blocks[[1]]$cumulative.distance)
old_path_no_blocks
```
```{r}
names(old_path[[1]])[1:2] <- c("a", "b")
names(old_path_no_blocks[[1]])[1:2] <- c("a", "b")
plotMatrix(
  distance.matrix = old_dist_matrix,
  least.cost.path = old_path_no_blocks
)
```


## Cost sum

### No blocks path



```{r}
cost_sum <- leastCost(
  least.cost.path = old_path_no_blocks
  )
cost_sum[[1]]
```

### Full path

Both versions are identical here

```{r}
old_cost_sum <- leastCost(
  least.cost.path = old_path
  )
old_cost_sum[[1]]
```

## Autosum

### no blocks path

```{r}
sequence_sum <- autoSum(
  sequences = sequences,
  least.cost.path = old_path_no_blocks,
  method = "euclidean"
  )
sequence_sum[[1]]
```

### Full path

Both versions return identical results

```{r}
colnames(old_path[[1]])[1:2] <- c("A", "B")
old_sequence_sum <- autoSum(
  sequences = sequences,
  least.cost.path = old_path,
  method = "euclidean"
  )
old_sequence_sum
```

## Psi

Both versions yield the same result

```{r}
old_dissimilarity <- psi(
  least.cost = old_cost_sum,
  autosum = old_sequence_sum
  )
old_dissimilarity[[1]]
```

## Complete psi computation

### With diagonals

Both versions yield the same results

```{r}
old_dissimilarity <- workflowPsi(
  sequences = sequences,
  grouping.column = "id",
  method = "euclidean",
  format = "list",
  diagonal = TRUE,
  ignore.blocks = FALSE
)
old_dissimilarity[[1]]
```

## P-value

```{r}
old_p_value <- workflowNullPsi(
  sequences = sequences,
  grouping.column = "id",
  method = "euclidean",
  diagonal = FALSE,
  ignore.blocks = FALSE,
  repetitions = 99,
  parallel.execution = TRUE
)
old_p_value$p
```

### Distribution

```{r}
old_null_values <- as.numeric(old_p_value$psi[1, 4:ncol(old_p_value$psi)])
old_null_values
```

## Importance

```{r}
importance <- workflowImportance(
  sequences = sequences,
  grouping.column = "id",
  method = "euclidean",
  diagonal = FALSE,
  ignore.blocks = TRUE
  )

importance$psi

importance$psi.drop
```

