---
title: "Mapping Time Series Dissimilarity"
output: 
  rmarkdown::html_document:
    toc: true
    toc_title: "Content"
    source: false
---

#WORK IN PROGRESS

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  dpi = 150,
  fig.show = 'hold',
  fig.width = 5
)

options(scipen = 9999)
options(conflicts.policy = list(warn = FALSE))
```

## Summary

This article focuses on the spatial representation of dissimilarity scores.

TODO

## Setup

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
if (!requireNamespace("sf", quietly = TRUE)){
  install.packages("sf")
}
if (!requireNamespace("mapview", quietly = TRUE)){
  install.packages("mapview")
}
if (!requireNamespace("dplyr", quietly = TRUE)){
  install.packages("dplyr")
}
```

```{r, message=FALSE, warning=FALSE, results="hide"}
library(distantia, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
library(mapview, warn.conflicts = FALSE)
```


## Example Data

The examples below illustrate how to map dissimilarity scores using the datasets `covid_prevalence` and `covid_polygons` included with `distantia`. The dataset comprises time series of weekly Covid-19 prevalence in several California counties.

```{r, fig.height=4.5}
tsl <- distantia::tsl_initialize(
  x = distantia::covid_prevalence,
  name_column = "name",
  time_column = "time"
)

distantia::tsl_plot(
  tsl = tsl[1:10],
  columns = 2,
  guide = FALSE
)
```
The map below displays the polygons in `distantia::covid_counties` using `mapview()`.

```{r}
mapview::mapview(
  distantia::covid_counties,
  label = "name"
)
```


## Dissimilarity Analysis

The code in this section prepares several datasets to use in the different mapping examples: 

  - `df_psi`: a lock-step dissimilarity data frame with *psi* scores and p-values.
  - `df_stats`: dissimilarity stats of each time series against all others.
  - `df_cluster`: a hierarchical clustering based on the dissimilarity scores in `df_psi`.

The lock-step dissimilarity analysis shown below includes p-values from a restricted permutation test. These p-values will be useful as criteria to select relevant mapping features.

```{r}
#parallelization setup
future::plan(
  future::multisession,
  workers = parallelly::availableCores() - 1
  )

#lock-step dissimilarity analysis
df_psi <- distantia::distantia(
  tsl = tsl,
  distance = "euclidean",
  lock_step = TRUE,
  repetitions = 1000,
  permutation = "restricted",
  block_size = 12 #weeks
)

#disable parallelization
future::plan(
  future::sequential
  )

#check resulting data frame
df_psi |> 
  dplyr::select(x, y, psi, p_value) |> 
  dplyr::glimpse()
```

The code below aggregates *psi* scores in the data frame `df_psi` to summarize the overall dissimilarity of each time series with all others.

```{r}
df_stats <- distantia::distantia_stats(
  df = df_psi
)

df_stats |> 
  dplyr::select(
    name, mean
  ) |> 
  dplyr::glimpse()
```

The function `distantia_cluster_hclust()` runs a hierarchical clustering using the *psi* scores in `df_psi` as clustering criteria.

```{r}
df_cluster <- distantia::distantia_cluster_hclust(
  df = df_psi
)$df

df_cluster |> 
  dplyr::select(name, cluster) |>
  dplyr::glimpse()
```

## Dissimilarity Network

The function `distantia_spatial_network()` transforms the result of `distantia()` to an sf data frame with edges connecting time series coordinates or polygons. The result can be interpreted as a *dissimilarity network*. 

```{r, message = FALSE, warning = FALSE}
sf_network <- distantia::distantia_spatial_network(
  df = df_psi,
  sf = distantia::covid_counties |> 
    dplyr::select(
      name, geometry
    )
)

dplyr::glimpse(sf_network)
```

The resulting sf data frame has a field with the edge name, the columns `x` and `y` with the names of the connected time series, the *psi* scores and p-values of the dissimilarity data frame, a geometry column of type LINESTRING defining the network edges, and the `length` of the edges.

This sf data frame can be mapped right away, but in this case there are too many pairs of counties to achieve a meaningful map

```{r, message = FALSE, warning = FALSE}
mapview::mapview(
  sf_network,
  layer.name = "Psi",
  label = "edge_name",
  zcol = "psi",
  color = distantia::color_continuous()
)
```

Focusing on particular aspects of the data at hand may help untangle this mess. For example, the code below subsets edges connecting with *San Francisco* and its most similar counties in terms of Covid19 prevalence.

```{r, warning = FALSE}
#subset edges involving San Francisco
#with a maximum length of 250km
sf_network_subset <- sf_network |> 
  dplyr::filter(
    grepl(
      pattern = "San_Francisco",
      x = edge_name
    ),
    p_value <= 0.05
  )

#map country polygons and dissimilarity edges
mapview::mapview(
  covid_counties,
  col.regions = NA,
  color = "black",
  label = "name",
  legend = FALSE,
  map.type = "OpenStreetMap"
) +
  mapview::mapview(
    sf_network_subset,
    layer.name = "San Francisco - Psi",
    label = "edge_name",
    zcol = "psi",
    lwd = 3,
    color = distantia::color_continuous(
      n = 5,
      rev = TRUE
      )
  )
```

A slightly more involved way to represent a one-to-many network with the data in `sf_network` and the polygons in `distantia::covid_counties` is shown below.

```{r, warning=FALSE}
sf_network_subset <- sf_network |> 
  sf::st_drop_geometry() |> 
  #filter edges not containing San_Francisco
  dplyr::filter(
    grepl(
      pattern = "San_Francisco",
      x = edge_name
    )
  ) |> 
  #create column "name" for join with covid_counties
  dplyr::mutate(
    name = ifelse(
      test = x != "San_Francisco", 
      yes = x, 
      no = y
    )
  ) |> 
  #add row with San Francisco
  dplyr::select(name, psi, p_value) |> 
  dplyr::bind_rows(
    data.frame(
      name = "San_Francisco",
      psi = 0,
      p_value = 0
    )
  ) |> 
  #join with polygons
  dplyr::inner_join(
    y = distantia::covid_counties |> 
      dplyr::select(name, geometry),
    by = "name"
  ) |> 
  #convert to sf data frame
  sf::st_sf()

#map result
mapview::mapview(
    sf_network_subset,
    layer.name = "Psi",
    label = "name",
    zcol = "psi",
    col.regions = distantia::color_continuous(
      rev = TRUE
      )
  )
```


## Dissimilarity Stats

Mapping the dissimilarity stats of each time series may help identify places that are somewhat *special* because they show a high dissimilarity with all others, or places that are *average* and have no distinctive features.

Merging the dissimilarity stats with the sf data frame containing the county polygons generates the spatial data required for the map.

```{r, warning = FALSE}
sf_stats <- merge(
  x = distantia::covid_counties,
  y = df_stats
) |> 
  dplyr::select(
    mean,
    name
  )
```

The map below uses warm colors to highlight places with a high dissimilarity with all others, such as Humboldt county in the north west of the state. 

```{r, warning = FALSE}
mapview::mapview(
  sf_stats,
  layer.name = "Psi mean",
  zcol = "mean",
  label = "name",
  col.regions = distantia::color_continuous(),
  alpha.regions = 1
)
```

## Dissimilarity Clustering

This section shows how to map the similarity groups resulting from `distantia::distantia_cluster_hclust()` or `distantia::distantia_cluster_kmeans()`.

The first block in the code below joins the county polygons in `distantia::covid_polygons` with the clustering data frame `df_cluster`, while creating a new variable "alpha" based on the column "silhouette_width", which represents the strength of group membership.

The second block generates the map, using the new variable to define transparency. This setup allows to identify each group of similar counties while helping identify these counties that somehow do not fully belong. This is the case of *Butte*, which is in the group 5, but with a very low silhouette score.

```{r}
sf_cluster <- distantia::covid_counties |> 
  dplyr::select(name, geometry) |> 
  dplyr::inner_join(
    y = df_cluster,
    by = "name"
  ) |> 
  dplyr::mutate(
    alpha = f_rescale_local(
      x = silhouette_width,
      new_min = 0.1
      )
  )

mapview::mapview(
  sf_cluster,
  layer.name = "Group",
  zcol = "cluster",
  label = "name",
  col.regions = distantia::color_continuous(
    n = max(sf_cluster$cluster)
  ),
  alpha.regions = sf_cluster$alpha
)
```

