---
title: "Mapping Time Series Dissimilarity"
output: 
  rmarkdown::html_document:
    toc: true
    toc_title: "Content"
    source: false
---

#WORK IN PROGRESS

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  dpi = 150,
  fig.show = 'hold',
  fig.width = 5
)

options(scipen = 9999)
options(conflicts.policy = list(warn = FALSE))
```

## Summary

Many time series happen at a given space, which can be represented either by a polygon, or by a pair of coordinates. The R package `distantia`, albeit focused on time series comparison, offers limited support to map time series dissimilarity.

This article illustrates this concept with several examples.

## Setup

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
  pak::pkg_install(
    c(
    "r-tmap/tmap"
    )
  )
}
```

```{r, message=FALSE, warning=FALSE}
library(distantia)
library(tmap)

#interactive maps 
tmap::tmap_mode("view")
```

## Mapping Dissimilarity Stats

Mapping the dissimilarity stats of each time series  helps identify places that are somewhat *special* because they show a high dissimilarity with all others, or places that are *average* and have no distinctive features.

The example below illustrates this concept using the datasets `covid_prevalence` and `covid_polygons` included with `distantia`.

The dataset comprises time series of weekly Covid19 prevalence in several California counties.

```{r, fig.height=4.5}
tsl <- distantia::tsl_initialize(
  x = distantia::covid_prevalence,
  name_column = "name",
  time_column = "time"
)

tsl_plot(
  tsl = tsl[1:10],
  columns = 2,
  guide = FALSE
)
```

In this case, the data requires no scaling because we will be computing lock-step dissimilarity to highlight differences in synchronicity and magnitude between prevalence curves.

```{r}
df_psi <- distantia::distantia(
  tsl = tsl,
  lock_step = TRUE
)
```

The code below computes the dissimilarity stats of each time series against all others.

```{r}
df_stats <- distantia::distantia_stats(
  df = df_psi
)

head(df_stats)
```

Merging the dissimilarity stats with the sf data frame containing the county polygons generates the spatial data required for the map.

```{r, warning = FALSE}
df_map <- merge(
  x = distantia::covid_counties,
  y = df_stats
)
```

The map below uses warm colors for places with a high dissimilarity with all others such as Humboldt county in the north west of the state, and cold colors for places with a higher dissimilarity with all others. 

```{r}
tmap::tm_shape(df_map) +
  tmap::tm_polygons(
    fill = "mean",
    fill.legend = tmap::tm_legend("Mean Psi"),
    fill.scale = tmap::tm_scale_continuous(
      values = distantia::utils_color_continuous_default(n = 5)
    )
  )
```

