---
title: "Mapping Time Series Dissimilarity"
output: 
  rmarkdown::html_document:
    toc: true
    toc_title: "Content"
    source: false
---

#WORK IN PROGRESS

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  dpi = 150,
  fig.show = 'hold',
  fig.width = 5
)

options(scipen = 9999)
options(conflicts.policy = list(warn = FALSE))
```

## Summary

Many time series happen at a given space, which can be represented either by a polygon, or by a pair of coordinates. The R package `distantia`, albeit focused on time series comparison, offers limited support to map time series dissimilarity.

This article illustrates this concept with several examples.

## Setup

```{r, echo = FALSE}
if(requireNamespace(package = "remotes") == FALSE){
  install.packages("remotes")
}
if(requireNamespace(package = "tmap") == FALSE){
  remotes::install_github("r-tmap/tmap")
}
if(requireNamespace(package = "dplyr") == FALSE){
  install.packages("dplyr")
}
```

```{r, message=FALSE, warning=FALSE}
library(distantia)
library(dplyr)
library(tmap)
library(future)
library(parallelly)

#interactive maps 
tmap_mode("view")

#parallelization setup
future::plan(
  future::multisession,
  workers = parallelly::availableCores() - 1
  )
```

## Covid

```{r, fig.height=4}
tsl <- distantia::tsl_initialize(
  x = distantia::covid_prevalence,
  name_column = "name",
  time_column = "time"
)

distantia::tsl_plot(
  tsl = tsl[1:4]
)
```

```{r}
df_psi <- distantia::distantia(
  tsl = tsl,
  lock_step = TRUE
)
```


```{r}
df_stats <- distantia::distantia_stats(
  df = df_psi
)

head(df_stats)
```


```{r, warning = FALSE}
df_map <- merge(
  x = distantia::covid_counties,
  y = df_stats
)
```

Mean average dissimilarity with all other counties.

```{r}
tmap::tm_shape(df_map) +
  tmap::tm_polygons(
    fill = "mean",
    fill.legend = tmap::tm_legend("Mean Psi"),
    fill.scale = tmap::tm_scale_continuous(
      values = viridis::turbo(n = 100))
    )
```

