---
title: "The Psi Dissimilarity Metric"
output: 
  rmarkdown::html_document:
    toc: true
    toc_title: "Content"
    source: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.show = 'hold' #side by side images
)
```

# Summary

This article builds upon [Benito and Birks 2020](https://doi.org/10.1111/ecog.04895) to describe in detail the computation of the dissimilarity metric used in `distantia` to compare pairs of time series. This content aims to provide a deep understanding on how dissimilarity is computed for different types of time series and purposes by mixing theoretical details with practical examples.

# The Psi Dissimilarity Metric

The Psi Dissimilarity Metric ($\psi$ hereafter) measures the distance between two time series in the range $[0, \infty)$, where 0 represents identical time series. 

In essence, $\psi$ is two times the sum of distances between all *relevant* pairs of samples of two time series, normalized by the sum of distances between consecutive samples in these time series.

The general expression to compute $\psi$ can be simplified to:

$$\psi = \frac{2 \times D(X, Y)}{d(X) + d(Y)} - 1$$

where:

  + $X$ and $Y$: time series matrices with the same number of columns.
  + $D$: function to sum distances between pairs of rows from each matrix.
  + $d$: function to sum distances between consecutive rows in a matrix.
  + $2 \times D(X, Y)$: distance between time series (named $AB_{between}$ in [Benito and Birks 2020](https://doi.org/10.1111/ecog.04895)).
  + $d(X) + d(Y)$: normalization factor (named $AB_{within}$ in the original paper).

```{r}
library(distantia)

X <- zoo::zoo(x = as.matrix(1:3), order.by = 1:3)
Y <- zoo::zoo(x = as.matrix(2:4), order.by = 1:3)

#normalization factor
########################

#base R
ab_within <- sum(diff(X)) + sum(diff(Y))

#distantia
ab_within <- auto_sum_full_cpp(X, Y)

#distance between time series
#############################

#base R
ab_between <- abs(sum(X - Y))*2

ab_between <- distance_lock_step_cpp(X, Y)


(ab_between - auto_sum) / auto_sum

(ab_between/auto_sum) - 1

psi(
  ab_between,
  ab_within,
  diagonal = FALSE
)
```



```{r setup}
library(distantia)
```




Going deeper into the math notation, $AB_{between}$ is expressed as:

$$AB_{\text{between}} = 2 \times \sum_{i=1}^{m=n} D(A_i, B_i)$$
where $m$ and $n$ are the number of samples in $A$ and $B$, $i$ is the relevant sample index, $n$ is the total number of sample pairs, and $D$ is a distance function (e.g., Euclidean, Manhattan, or any other relevant distance metric for the values in $A$ and $B$).

On the other hand, $AB_{within}$ is represented as:

$$AB_{\text{within}} = \sum_{i=2}^{m} D(A_{i-1}, A_i) + \sum_{i=2}^{n} D(B_{i-1}, B_i) $$
where $D(A_{i-1}, A_i)$ and $D(B_{i-1}, B_i)$ represent the sum of distances between consecutive samples within $A$ and $B$, respectively.
