---
title: "Handling Time Series Lists"
output: 
  rmarkdown::html_document:
    toc: true
    toc_title: "Content"
    source: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  fig.show = 'hold',
  fig.width = 5
)
```


## Summary

The R package `distantia` 2.0 introduces Time Series Lists as working unit for dissimilarity analyses, and provides a complete toolset to manages them.

This article describes Time Series Lists in detail, and showcases the most common data handling procedures enabled by the new functions included in the package. 

## Introduction to Time Series Lists

In this new version of `distantia`, groups of time series are organized as named lists of [zoo](https://cran.r-project.org/web/packages/zoo/index.html) objects. These lists are named **Time Series Lists** (TSL) within the package, and are designed to facilitate the parallelization of dissimilarity analyses. 

TSL is not a class by choice, as the idea is keeping them as simple as possible to grant users the autonomy to create or modify them as needed.

### Zoo Time Series

The R package `zoo` provides an S3 class of the same name designed to handle observations ordered by an index. It supports various index classes, such as Date, POSIXct, or even custom numeric or character indices, and handles regular and irregular time series equally well.

Other advantages of using zoo objects include a seamless intergration with base R methods, and built-in tools for alignment, merging, and subsetting.

Let's take a look at a little zoo object.

```{r, fig.width=5, fig.height=3}
library(distantia)

z <- zoo_simulate(
  name = "my_zoo",
  cols = 3,
  rows = 10,
  time_range = c(
    "2024-01-01", 
    "2024-12-31"
    ),
  na_fraction = 0.1,
  irregular = TRUE,
  seed = 1
)

zoo_plot(x = z)
```

Zoo objects have two main components, a data matrix with the time series observations, and an index representing time or sample order.

The data matrix is extracted with `zoo::coredata()`.

```{r}
zoo::coredata(z)
class(zoo::coredata(z))
```
The core data of a zoo object can also be a vector when the time series is univariate.

```{r}
x <- zoo::zoo(x = runif(10))
is.vector(zoo::coredata(x))
```
However, this is **frowned upon** in `distantia`, and these vectors should always be converted to matrices.

```{r}
x <- distantia::zoo_vector_to_matrix(x = x)
is.vector(zoo::coredata(x))
is.matrix(zoo::coredata(x))
```

The index of zoo time series is extracted with `zoo::index()`.

```{r}
zoo::index(z)
class(zoo::index(z))
```
The classes for zoo indices explicitly supported in `distantia` are `Date`, `POSIXct`, and `numeric`. The function `distantia::zoo_time()` helps summarize the time features of a zoo object

```{r}
distantia::zoo_time(x = z)
```

Additionally, in `distantia` all zoo objects are expected to have the attribute `name`.

```{r}
attributes(z)$name
```
This attribute is used to facilitate plotting operations, and it is managed internally by `tsl_...()` functions. There are several functions in `distantia` to manage the names of zoo objects.

```{r}
#reset zoo name
z <- distantia::zoo_name_set(
  x = z,
  name = "My_Zoo"
)

#get zoo name
distantia::zoo_name_get(x = z)

#clean zoo name
z <- distantia::zoo_name_clean(
  x = z,
  lowercase = TRUE
)

distantia::zoo_name_get(x = z)
```
This package comes with several functions to manipulate zoo objects:

  - `zoo_name_set()`, `zoo_name_get()` and `zoo_name_clean()`: handle the attribute "name".
  - `zoo_time()`: details of the zoo index.
  - `zoo_aggregate()`: time aggregation of zoo objects.
  - `zoo_resample()`: interpolation or extrapolation to a different time, or from irregular to regular time.
  - `zoo_permute()`: restricted permutation.
  - `zoo_vector_to_matrix()` and `zoo_to_tsl()`: internal functions to facilitate handling zoo objects within TSLs.

### Time Series Lists (TSL)

TSLs are *named* lists of zoo time series. The example below shows how to build a TSL from scratch with zoo objects. But this is not the most common or comfortable case, so please, visit the section **Creating Time Series Lists** to find out how to convert your data easily to TSL.

```{r}
#create simple tsl
my_tsl <- list(
  A = zoo_simulate(
    cols = 4,
    na_fraction = 0.2
  ),
  B = zoo_simulate()
)

names(my_tsl)

class(my_tsl)

#names of the zoo objects
lapply(X = my_tsl, FUN = distantia::zoo_name_get)

#class of the objects within the list
lapply(X = my_tsl, FUN = class)
```

TSLs ready for dissimilarity analyses **must follow several rules**:

  - The elements of the list or the zoo objects must be named.
  - Time series names cannot be duplicated.
  - The coredata of all zoo objects must be of class `matrix`.
  - There cannot be NA, Inf, or NaN in the zoo objects.
  - All zoo columns must be named.
  - All zoo columns must be numeric.
  - All zoo objects within the TSL must share at least one column name.
  - The index of the zoo objects must be of the same class.
  
These are many rules, I know, but the functions `tsl_diagnose()` and `tsl_repair()` are there to help you forget about them.

When applying `tsl_diagnose()` to `my_tsl` we can see it has several issues.

```{r}
distantia::tsl_diagnose(tsl = my_tsl)
```
From there, we can either follow the suggestions, or apply `tsl_repair()` directly, as done below.

```{r}
my_tsl <- distantia::tsl_repair(tsl = my_tsl)
```
This function identifies the issues raised up by `tsl_diagnose()` and repairs them when possible.

If we run `tsl_diagnose()` again, all checks should be ok.

```{r}
distantia::tsl_diagnose(
  tsl = my_tsl
)
```

From this point, our TSL is ready to go!

```{r}
distantia::tsl_plot(
  tsl = my_tsl
)
```

## Creating TSLs

The function `tsl_initialize()` (with the alias `tsl_init()`) is designed to help transform several data structures to Time Series List.

### Long Data Frame to TSL

Long and tidy data frames are convenient structures to store multivariate time series of a reasonable size. For example, the data frame `fagus_dynamics` shown below has the column "name" identifying separate time series, the column "time" with observation dates, and three numeric columns with environmental observations.

```{r}
head(fagus_dynamics)
```
Transforming this data frame to TSL is quite straightforward:

```{r}
tsl <- tsl_initialize(
  x = fagus_dynamics,
  name_column = "name",
  time_column = "time"
)

#even shorter!
tsl <- tsl_init(
  x = fagus_dynamics,
  name = "name",
  time = "time"
)

tsl_plot(
  tsl  = tsl
)
```
Once manipulated and/or analyzed, a TSL can be converted back to data frame with `tsl_to_df()`.

```{r}
df <- tsl_to_df(tsl = tsl)
head(df)
```

### Wide Data Frame to TSL

```{r, echo = FALSE}
evi_wide <- stats::reshape(
  data = fagus_dynamics[, c(
    "name",
    "time",
    "evi"
  )],
  timevar = "name",
  idvar = "time",
  direction = "wide",
  sep = "_"
)
```

A wide data frame is a useful structure to store univariate time series observed in different places at the same times.

```{r}
head(evi_wide)
```
When no `name_column` is provided, `tsl_initialize()` assumes that the time series are coded as separate columns.

```{r}
tsl <- tsl_initialize(
  x = evi_wide,
  time_column = "time"
  )

tsl_plot(
  tsl = tsl
)
```
In this case, the column names of the univariate zoo objects will have the same name ("x").

```{r}
tsl_colnames_get(tsl = tsl)
```
This name can be reset as needed with `tsl_colnames_set()`.

```{r}
tsl <- tsl_colnames_set(
  tsl = tsl,
  names = "evi"
)

tsl_colnames_get(tsl = tsl)
```
This TSL can be converted to data frame, but this time the result comes in long format.

```{r}
df <- tsl_to_df(tsl = tsl)
head(df)
```

### Vectors and Matrices to TSL

A list of numeric vectors can also be converted to TSL. In this case, the zoo index is a sequence of integers.

```{r, fig.height=3.5}
tsl <- tsl_initialize(
  x = list(
    a = runif(10),
    b = runif(10)
  )
)

tsl_plot(
  tsl = tsl
)
```

The same thing can be done with matrices as well.

```{r, fig.height=3.5}
tsl <- tsl_initialize(
  x = list(
    a = matrix(data = runif(100), ncol = 2, nrow = 50),
    b = matrix(data = runif(100), ncol = 2, nrow = 50)
  )
)

tsl_plot(
  tsl = tsl
)
```
## The TSL Toolset

All functions in `distantia` to handle TSLs are named `tsl_...()`. Several of these functions are designed to explore and better understand your TSLs, while others are designed to manipulate and transform them to facilitate dissimilarity analyses.

### Parallelization Setup

All TSL functions have something in common: **they are a bunch of `lapply` in a trench coat**. Most use a parallelized version of `lapply` from the `future.apply` package, `future.apply::future_lapply()`, and a few combine `foreach::foreach()` with `doFuture::`%dofuture%` in parallelized loops.

As such, they all support a parallelization backend provided by the `future` package, as shown below.

```{r}
library(future)
library(parallelly)

future::plan(
  future::multisession,
  workers = parallelly::availableCores() - 1
  )

#progress bar (does not work in Rmarkdown)
#progressr::handlers(global = TRUE)
```

They also support progress bars via the [`progressr`](https://progressr.futureverse.org/) package. However, this option, commented in the code above, does not work in Rmarkdown.

### Explore TSLs

This section showcases the tools available in `distantia` to extract details from our Time Series Lists. But first, let's create a simulated TSL.

```{r, fig.height=6}
tsl <- tsl_simulate(
  n = 4,
  rows = 1000,
  seasons = 10,
  na_fraction = 0.1,
  irregular = TRUE
)

tsl_plot(tsl = tsl)
```

The most general tools in this section are focused on simple things such as TSL names and dimensions.

```{r}
#time series names
tsl_names_get(tsl = tsl)

#time series column names
tsl_colnames_get(tsl = tsl)

#number of columns
tsl_ncol(tsl = tsl)

#number of roes
tsl_nrow(tsl = tsl)
```

The function `tsl_time()` analyzes the time features of all time series in a TSL and generates a data frame.

```{r}
tsl_time(tsl = tsl)
```
```{r}
tsl_time
```


The function `tsl_stats()` focuses on the data values instead to provides a summary of their stats by time series and variable.

```{r}
df_stats <- tsl_stats(
  tsl = tsl, 
  lags = 1:2 #temporal autocorrelation lags
  )

df_stats
```
### Handle NA Cases

Dissimilarity analyses with `distantia`, because they designed with computational efficiency in mind, **do not support NA cases**. 

There are two alternate workflows to handle NA data in time series list.

The first uses `tsl_count_NA()` and `tsl_handle_NA()`. The former converts Inf and NaN to NA and counts NA cases in each time series, while the latter either omits or imputes NA cases via `zoo::na.approx()`.

```{r}
tsl_count_NA(tsl = tsl) 

tsl_notNA <- tsl_handle_NA(
  tsl = tsl,
  na_action = "impute"
)

tsl_count_NA(tsl = tsl_notNA) 
```
The second workflow is more general because it can address other potential issues. It involves using `tsl_diagnose()` and `tsl_repair()`.

```{r}
#diagnose issues with NA values
tsl_diagnose(tsl = tsl)

#impute NA cases
tsl <- tsl_repair(tsl = tsl)

#re-diagnose to check result
tsl_diagnose(tsl = tsl)
```

### Subsetting

The function `tsl_subset()` helps focus on particular *regions* of a time series list. Additionally, this function returns by default the numeric columns that are shared across all time series in a TSL.

```{r, fig.height=4.5}
#check details
tsl_names_get(tsl = tsl)
tsl_colnames_get(tsl = tsl)
tsl_time(tsl = tsl)[, c("name", "begin", "end")]

#subset
tsl_new <- tsl_subset(
  tsl = tsl,
  names = c("A", "C", "D"),
  colnames = c("a", "b"),
  time = c("2014-01-01", "2018-01-01")
)

#check details of new tsl
tsl_names_get(tsl = tsl_new)
tsl_colnames_get(tsl = tsl_new)
tsl_time(tsl = tsl_new)[, c("name", "begin", "end")]

#plotting the new tsl
tsl_plot(tsl = tsl_new)
```

The function `tsl_burst()` transforms a multivariate TSL into a univariate TSL by creating a new zoo object from each column of the original zoo objects. This function helps apply dissimilarity analysis between individual variables in multivariate time series.


```{r, fig.height=5}
tsl_univariate <- tsl_burst(
  tsl = tsl_new
)

#check new names and colnames
tsl_names_get(tsl = tsl)
tsl_colnames_get(tsl = tsl)

#plot new tsl
tsl_plot(tsl = tsl_univariate)

```

